---
title: "Data Cleaning Report"
date: last-modified
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    self-contained: true
    code-tools: false
    smooth-scroll: true
params:
  work_dir: null
  objects: null
  metadata: null
  cleaning_params: null
editor:
  markdown:
    wrap: sentence
---

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  fig.width = 10,
  fig.height = 6
)

library(massdataset)
library(massqc)
library(ggplot2)
library(dplyr)
library(tidyr)
library(plotly)

`%||%` <- function(x, y) if (is.null(x) || length(x) == 0 || all(is.na(x))) y else x

work_dir <- params$work_dir %||% "."

step_info <- data.frame(
  step_id = c("raw", "clean", "outlier", "impute", "norm"),
  step_label = c(
    "Raw peak table",
    "Noise removal",
    "Outlier detection",
    "Missing value imputation",
    "Normalization"
  ),
  pos_name = c(
    "object_pos_raw",
    "object_pos_clean",
    "object_pos_outlier",
    "object_pos_impute",
    "object_pos_norm"
  ),
  neg_name = c(
    "object_neg_raw",
    "object_neg_clean",
    "object_neg_outlier",
    "object_neg_impute",
    "object_neg_norm"
  ),
  stringsAsFactors = FALSE
)

candidate_paths <- function(mode, step_id, object_name) {
  mode <- tolower(mode)
  aliases <- c(
    paste0(object_name, ".rds"),
    paste0(object_name, ".RDS"),
    paste0(object_name, ".rda"),
    paste0(object_name, ".RData")
  )
  legacy <- switch(
    paste(mode, step_id, sep = "_"),
    "pos_raw" = c("object_pos.rds", "object_pos_raw.rds", "01.object_pos_raw.rda"),
    "neg_raw" = c("object_neg.rds", "object_neg_raw.rds", "01.object_neg_raw.rda"),
    "pos_clean" = c("object_pos_clean.rds", "object_pos_mv.rds", "02.object_pos_mv.rda"),
    "neg_clean" = c("object_neg_clean.rds", "object_neg_mv.rds", "02.object_neg_mv.rda"),
    "pos_outlier" = c("object_pos_outlier.rds", "03.object_pos_outlier.rda"),
    "neg_outlier" = c("object_neg_outlier.rds", "03.object_neg_outlier.rda"),
    "pos_impute" = c("object_pos_impute.rds", "04.object_pos_impute.rda"),
    "neg_impute" = c("object_neg_impute.rds", "04.object_neg_impute.rda"),
    "pos_norm" = c("object_pos_norm.rds", "05.object_pos_norm.rda"),
    "neg_norm" = c("object_neg_norm.rds", "05.object_neg_norm.rda"),
    character()
  )
  candidates <- c(aliases, legacy)
  unique(c(
    file.path(work_dir, candidates),
    file.path(work_dir, "objects", candidates)
  ))
}

load_mass_object <- function(paths, object_name = NULL) {
  paths <- paths[file.exists(paths)]
  if (length(paths) == 0) return(NULL)
  path <- paths[1]
  ext <- tolower(tools::file_ext(path))
  if (ext == "rds") {
    obj <- readRDS(path)
    if (inherits(obj, "mass_dataset")) return(obj)
    return(NULL)
  }
  e <- new.env(parent = emptyenv())
  load(path, envir = e)
  if (!is.null(object_name) && exists(object_name, envir = e, inherits = FALSE)) {
    obj <- get(object_name, envir = e)
    if (inherits(obj, "mass_dataset")) return(obj)
  }
  objects <- mget(ls(e), envir = e)
  hits <- objects[vapply(objects, inherits, logical(1), what = "mass_dataset")]
  if (length(hits) == 0) return(NULL)
  hits[[1]]
}

object_from_param <- function(mode, step_id) {
  obj_map <- params$objects
  if (is.null(obj_map)) return(NULL)
  key <- paste(tolower(mode), step_id, sep = "_")
  value <- obj_map[[key]] %||% obj_map[[toupper(mode)]][[step_id]] %||% NULL
  if (is.null(value)) return(NULL)
  if (inherits(value, "mass_dataset")) return(value)
  if (is.character(value)) return(load_mass_object(value))
  NULL
}

objects_by_step <- lapply(seq_len(nrow(step_info)), function(i) {
  step_id <- step_info$step_id[[i]]
  list(
    POS = object_from_param("pos", step_id) %||%
      load_mass_object(candidate_paths("pos", step_id, step_info$pos_name[[i]]), step_info$pos_name[[i]]),
    NEG = object_from_param("neg", step_id) %||%
      load_mass_object(candidate_paths("neg", step_id, step_info$neg_name[[i]]), step_info$neg_name[[i]])
  )
})
names(objects_by_step) <- step_info$step_id

available_object <- function(mode = c("POS", "NEG")) {
  mode <- match.arg(mode)
  for (step_id in rev(names(objects_by_step))) {
    obj <- objects_by_step[[step_id]][[mode]]
    if (!is.null(obj)) return(obj)
  }
  NULL
}

object_status <- function(obj) {
  if (is.null(obj)) {
    return(data.frame(
      samples = NA_integer_,
      variables = NA_integer_,
      missing_values = NA_integer_,
      missing_fraction = NA_real_,
      zero_values = NA_integer_,
      zero_fraction = NA_real_,
      qc_samples = NA_integer_,
      biological_samples = NA_integer_
    ))
  }
  expr <- as.matrix(massdataset::extract_expression_data(obj))
  sample_info <- as.data.frame(massdataset::extract_sample_info(obj), stringsAsFactors = FALSE)
  class_col <- first_existing_col(sample_info, c("class", "sample_type", "sample_class"))
  sample_class <- if (!is.null(class_col)) tolower(as.character(sample_info[[class_col]])) else rep("", nrow(sample_info))
  data.frame(
    samples = ncol(expr),
    variables = nrow(expr),
    missing_values = sum(is.na(expr)),
    missing_fraction = mean(is.na(expr)),
    zero_values = sum(expr == 0, na.rm = TRUE),
    zero_fraction = mean(expr == 0, na.rm = TRUE),
    qc_samples = sum(sample_class == "qc", na.rm = TRUE),
    biological_samples = sum(!sample_class %in% c("qc", "blank", ""), na.rm = TRUE)
  )
}

first_existing_col <- function(df, candidates) {
  hit <- candidates[candidates %in% colnames(df)]
  if (length(hit) == 0) return(NULL)
  hit[1]
}

plot_color_column <- function(obj) {
  if (is.null(obj)) return(NULL)
  sample_info <- as.data.frame(massdataset::extract_sample_info(obj), stringsAsFactors = FALSE)
  if ("group" %in% colnames(sample_info)) {
    n_group <- length(unique(stats::na.omit(as.character(sample_info$group))))
    if (is.finite(n_group) && n_group <= 15) return("group")
  }
  first_existing_col(sample_info, c("class", "sample_type", "sample_class", "group", "batch", "sample_id"))
}

sample_order_column <- function(obj) {
  if (is.null(obj)) return("sample_id")
  sample_info <- as.data.frame(massdataset::extract_sample_info(obj), stringsAsFactors = FALSE)
  first_existing_col(sample_info, c("injection.order", "injection_order", "sample_id")) %||% "sample_id"
}

sample_count <- function(obj) {
  if (is.null(obj)) return(0L)
  ncol(as.matrix(massdataset::extract_expression_data(obj)))
}

interactive_plot <- function(obj) {
  sample_count(obj) <= 1000
}

render_report_plot <- function(plot, obj) {
  if (is.null(plot)) return(invisible(NULL))
  if (interactive_plot(obj)) {
    return(plotly::ggplotly(plot) |>
      plotly::layout(legend = list(orientation = "v")))
  } else {
    return(plot)
  }
}

plot_mode_label <- function(obj) {
  if (interactive_plot(obj)) {
    paste0("Interactive plotly rendering is enabled because this object has ", sample_count(obj), " samples (threshold <= 1000).")
  } else {
    paste0("Static rendering is used because this object has ", sample_count(obj), " samples (threshold > 1000).")
  }
}

safe_log_scale <- function(obj, scale = TRUE) {
  if (is.null(obj)) return(NULL)
  temp <- log(obj + 1, 2)
  if (scale) {
    ed <- temp@expression_data
    scaled <- t(scale(t(ed)))
    scaled[!is.finite(scaled)] <- 0
    temp@expression_data <- as.data.frame(scaled)
  }
  temp
}

status_table <- do.call(rbind, lapply(seq_len(nrow(step_info)), function(i) {
  step_id <- step_info$step_id[[i]]
  step_label <- step_info$step_label[[i]]
  do.call(rbind, lapply(c("POS", "NEG"), function(mode) {
    data.frame(
      step = step_label,
      step_id = step_id,
      mode = mode,
      object_status(objects_by_step[[step_id]][[mode]]),
      row.names = NULL
    )
  }))
}))

status_table <- status_table |>
  group_by(mode) |>
  mutate(
    variables_removed_from_previous = lag(variables) - variables,
    samples_removed_from_previous = lag(samples) - samples,
    missing_fraction_delta = missing_fraction - lag(missing_fraction)
  ) |>
  ungroup()

fmt_pct <- function(x) {
  ifelse(is.na(x), NA_character_, sprintf("%.2f%%", x * 100))
}

fmt_int <- function(x) {
  ifelse(is.na(x), NA_character_, format(as.integer(round(x)), big.mark = ","))
}

fmt_method <- function(x) {
  if (is.null(x) || length(x) == 0 || is.na(x[1]) || !nzchar(as.character(x[1]))) return("not specified")
  as.character(x[1])
}

get_param <- function(section, key, default = "not specified") {
  values <- cleaning_params[[section]]
  if (is.null(values) || is.null(values[[key]])) return(default)
  fmt_method(values[[key]])
}

sample_ids <- function(obj) {
  if (is.null(obj)) return(character())
  info <- as.data.frame(massdataset::extract_sample_info(obj), stringsAsFactors = FALSE)
  if (!"sample_id" %in% colnames(info)) return(character())
  as.character(info$sample_id)
}

removed_samples <- function(mode) {
  before <- sample_ids(objects_by_step$clean[[mode]])
  after <- sample_ids(objects_by_step$outlier[[mode]])
  setdiff(before, after)
}

row_for <- function(step_id, mode) {
  hit <- which(as.character(status_table$step_id) == as.character(step_id) &
                 as.character(status_table$mode) == as.character(mode))
  if (length(hit) == 0) return(status_table[0, , drop = FALSE])
  status_table[hit[1], , drop = FALSE]
}

status_value <- function(step_id, mode, col) {
  row <- row_for(step_id, mode)
  if (nrow(row) == 0 || !col %in% colnames(row)) return(NA)
  row[[col]][[1]]
}

highlight_numbers <- function(items) {
  paste(sprintf("**%s**", items), collapse = ", ")
}

method_label_outlier <- function() {
  method <- tolower(get_param("outlier_detection", "outlier_method", "auto"))
  if (method %in% c("auto", "tidymass", "by tidymass")) {
    paste0("TidyMass automatic detection", " (criteria: ", get_param("outlier_detection", "auto_criteria", "any"), ")")
  } else if (method %in% c("manual", "by myself")) {
    "manual selection"
  } else {
    method
  }
}

integration_label <- function() {
  obj <- objects_by_step$norm$POS %||% objects_by_step$norm$NEG
  integ <- get_param("normalization", "integ_method", "qc_mean")
  if (is.null(obj)) return(paste("batch integration method was set to", integ))
  info <- as.data.frame(massdataset::extract_sample_info(obj), stringsAsFactors = FALSE)
  if (!"batch" %in% colnames(info) || length(unique(stats::na.omit(as.character(info$batch)))) <= 1) {
    return("batch integration was skipped because only one batch was detected")
  }
  paste("batch integration used", integ)
}

callout <- function(title, text) {
  cat("::: {.callout-important appearance=\"simple\" icon=\"false\"}\n")
  cat("## ", title, "\n\n", sep = "")
  cat(text, "\n")
  cat(":::\n")
}

display_status <- status_table |>
  mutate(
    missing_fraction = fmt_pct(missing_fraction),
    zero_fraction = fmt_pct(zero_fraction),
    missing_fraction_delta = fmt_pct(missing_fraction_delta)
  )

cleaning_params <- params$cleaning_params %||% list()
```

::: {.callout-note icon="false" appearance="simple"}
## Report Information

**Software**: [MetMiner by Wang et al., 2024](https://github.com/ShawnWx2019/MetMiner), powered by [TidyMass2 by Wang et al., 2026](https://www.nature.com/articles/s41467-026-68464-7)

**Date**: `r format(Sys.Date(), '%Y-%m-%d')`

**Generated By**: [MetMiner-shinyapp](https://github.com/ShawnWx2019/MetMiner)
:::

------------------------------------------------------------------------

# Introduction

This report summarizes the step-by-step data cleaning workflow after peak picking.
It tracks how the `mass_dataset` object changes through noise removal, sample outlier handling, missing-value imputation, and normalization.

The PCA and sample-missing-value plots automatically choose the visual grouping column:
when the number of `group` levels is 15 or fewer, `group` is used; otherwise the report falls back to `class` or an equivalent sample class column.

------------------------------------------------------------------------

# Workflow Summary

```{r workflow-summary}
knitr::kable(
  display_status |>
    select(
      mode, step, samples, variables, missing_values, missing_fraction,
      zero_values, zero_fraction, qc_samples, biological_samples,
      variables_removed_from_previous, samples_removed_from_previous,
      missing_fraction_delta
    ),
  caption = "Mass dataset status at each data-cleaning step."
)
```

```{r}
#| label: workflow-change-plot
#| fig-height: 5
status_plot_data <- status_table |>
  filter(!is.na(variables)) |>
  mutate(step = factor(step, levels = step_info$step_label))

p <- ggplot(status_plot_data, aes(step, variables, group = mode, color = mode)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2.5) +
  labs(x = NULL, y = "Feature number", title = "Feature Retention Across Data Cleaning") +
  theme_light() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
p
```

```{r}
#| label: missing-change-plot
#| fig-height: 5
p <- ggplot(status_plot_data, aes(step, missing_fraction * 100, group = mode, color = mode)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2.5) +
  labs(x = NULL, y = "Missing values (%)", title = "Missing-Value Fraction Across Data Cleaning") +
  theme_light() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
p
```

::: {.callout-tip title="Figure Legend" collapse="true"}
These overview plots track the number of retained features and the missing-value fraction across the full cleaning workflow.

- `r plot_mode_label(available_object("POS") %||% available_object("NEG"))`
- A steep feature decrease is expected during noise removal; missing values should approach zero after imputation.
:::

------------------------------------------------------------------------

# Step-by-Step Diagnostics

## Raw Peak Table

::: panel-tabset
### POS

```{r}
#| label: raw-pos-plots
#| fig-height: 6
obj <- objects_by_step$raw$POS
if (!is.null(obj)) {
  color_col <- plot_color_column(obj)
  order_col <- sample_order_column(obj)
  p1 <- massdataset::show_sample_missing_values(obj, color_by = color_col, order_by = order_col, percentage = TRUE) +
    ggtitle(paste0("POS Raw Missing Values (colored by ", color_col, ")")) +
    theme_light() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
  render_report_plot(p1, obj)
}
```

```{r}
#| label: raw-pos-pca
#| fig-height: 6
obj <- objects_by_step$raw$POS
if (!is.null(obj)) {
  color_col <- plot_color_column(obj)
  p <- massqc::massqc_pca(safe_log_scale(obj, scale = TRUE), color_by = color_col, order_by = "sample_id") +
    ggtitle(paste0("POS Raw PCA (colored by ", color_col, ")")) +
    theme_light()
  render_report_plot(p, obj)
}
```

### NEG

```{r}
#| label: raw-neg-plots
#| fig-height: 6
obj <- objects_by_step$raw$NEG
if (!is.null(obj)) {
  color_col <- plot_color_column(obj)
  order_col <- sample_order_column(obj)
  p1 <- massdataset::show_sample_missing_values(obj, color_by = color_col, order_by = order_col, percentage = TRUE) +
    ggtitle(paste0("NEG Raw Missing Values (colored by ", color_col, ")")) +
    theme_light() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
  render_report_plot(p1, obj)
}
```

```{r}
#| label: raw-neg-pca
#| fig-height: 6
obj <- objects_by_step$raw$NEG
if (!is.null(obj)) {
  color_col <- plot_color_column(obj)
  p <- massqc::massqc_pca(safe_log_scale(obj, scale = TRUE), color_by = color_col, order_by = "sample_id") +
    ggtitle(paste0("NEG Raw PCA (colored by ", color_col, ")")) +
    theme_light()
  render_report_plot(p, obj)
}
```
:::

::: {.callout-tip title="Figure Legend" collapse="true"}
Raw peak-table plots show sample-level missing-value structure and PCA separation before data cleaning.

- **Color**: automatically uses `group` when there are 15 or fewer groups; otherwise it falls back to `class`.
- `r plot_mode_label(objects_by_step$raw$POS %||% objects_by_step$raw$NEG)`
- QC samples should cluster tightly in PCA and should not show unexpectedly high missing-value rates.
:::

## Noise Removal

```{r}
#| label: noise-removal-conclusion
#| results: asis
pos_removed <- status_value("clean", "POS", "variables_removed_from_previous")
neg_removed <- status_value("clean", "NEG", "variables_removed_from_previous")
pos_after <- status_value("clean", "POS", "variables")
neg_after <- status_value("clean", "NEG", "variables")
pos_mv <- status_value("clean", "POS", "missing_fraction")
neg_mv <- status_value("clean", "NEG", "missing_fraction")
method_bits <- c(
  paste0("missing-value filter: QC <= ", fmt_pct(as.numeric(get_param("remove_noise", "qc_na_freq", 0.2))),
         ", Subject <= ", fmt_pct(as.numeric(get_param("remove_noise", "s_na_freq", 0.5)))),
  paste0("QC RSD filter: ", get_param("remove_noise", "rm_noise_rsd", "TRUE"),
         " at ", get_param("remove_noise", "qc_rsd", "30"), "%"),
  paste0("intensity filter: ", get_param("remove_noise", "rm_intensity", "FALSE"))
)
callout(
  "Noise removal conclusion",
  paste0(
    "Noise removal used ", paste(method_bits, collapse = "; "), ". ",
    "It removed ", highlight_numbers(c(paste(fmt_int(pos_removed), "POS features"), paste(fmt_int(neg_removed), "NEG features"))),
    ", retaining ", highlight_numbers(c(paste(fmt_int(pos_after), "POS features"), paste(fmt_int(neg_after), "NEG features"))),
    ". The remaining missing-value fractions were ",
    highlight_numbers(c(paste0("POS ", fmt_pct(pos_mv)), paste0("NEG ", fmt_pct(neg_mv)))), "."
  )
)
```

```{r noise-removal-table}
noise_tbl <- status_table |>
  filter(step_id %in% c("raw", "clean")) |>
  select(mode, step, samples, variables, missing_fraction, variables_removed_from_previous, missing_fraction_delta)
knitr::kable(
  noise_tbl |>
    mutate(missing_fraction = fmt_pct(missing_fraction), missing_fraction_delta = fmt_pct(missing_fraction_delta)),
  caption = "Changes after noise removal."
)
```

::: panel-tabset
### POS

```{r}
#| label: clean-pos-missing
#| fig-height: 6
obj <- objects_by_step$clean$POS
if (!is.null(obj)) {
  color_col <- plot_color_column(obj)
  order_col <- sample_order_column(obj)
  p <- massdataset::show_sample_missing_values(obj, color_by = color_col, order_by = order_col, percentage = TRUE) +
    ggtitle(paste0("POS After Noise Removal (colored by ", color_col, ")")) +
    theme_light() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
  render_report_plot(p, obj)
}
```

### NEG

```{r}
#| label: clean-neg-missing
#| fig-height: 6
obj <- objects_by_step$clean$NEG
if (!is.null(obj)) {
  color_col <- plot_color_column(obj)
  order_col <- sample_order_column(obj)
  p <- massdataset::show_sample_missing_values(obj, color_by = color_col, order_by = order_col, percentage = TRUE) +
    ggtitle(paste0("NEG After Noise Removal (colored by ", color_col, ")")) +
    theme_light() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
  render_report_plot(p, obj)
}
```
:::

::: {.callout-tip title="Figure Legend" collapse="true"}
Noise-removal plots show the sample missing-value profile after feature filtering.

- **Color**: selected by the same group/class rule used throughout this report.
- `r plot_mode_label(objects_by_step$clean$POS %||% objects_by_step$clean$NEG)`
- Remaining high missing-value samples after this step may require outlier review.
:::

## Outlier Detection

```{r}
#| label: outlier-conclusion
#| results: asis
pos_removed_samples <- removed_samples("POS")
neg_removed_samples <- removed_samples("NEG")
pos_removed_n <- length(pos_removed_samples)
neg_removed_n <- length(neg_removed_samples)
sample_text <- function(ids) if (length(ids) == 0) "none" else paste(ids, collapse = ", ")
callout(
  "Outlier detection conclusion",
  paste0(
    "Outlier detection used ", highlight_numbers(method_label_outlier()), ". ",
    "It removed ", highlight_numbers(c(paste(pos_removed_n, "POS samples"), paste(neg_removed_n, "NEG samples"))), ". ",
    "Removed materials: POS ", highlight_numbers(sample_text(pos_removed_samples)),
    "; NEG ", highlight_numbers(sample_text(neg_removed_samples)), "."
  )
)
```

```{r outlier-table}
outlier_tbl <- status_table |>
  filter(step_id %in% c("clean", "outlier")) |>
  select(mode, step, samples, variables, samples_removed_from_previous, missing_fraction)
knitr::kable(
  outlier_tbl |>
    mutate(missing_fraction = fmt_pct(missing_fraction)),
  caption = "Changes after sample outlier detection."
)
```

::: panel-tabset
### POS

```{r}
#| label: outlier-pos-pca
#| fig-height: 6
obj <- objects_by_step$outlier$POS
if (!is.null(obj)) {
  color_col <- plot_color_column(obj)
  p <- massqc::massqc_pca(safe_log_scale(obj, scale = TRUE), color_by = color_col, order_by = "sample_id") +
    ggtitle(paste0("POS After Outlier Detection (colored by ", color_col, ")")) +
    theme_light()
  render_report_plot(p, obj)
}
```

### NEG

```{r}
#| label: outlier-neg-pca
#| fig-height: 6
obj <- objects_by_step$outlier$NEG
if (!is.null(obj)) {
  color_col <- plot_color_column(obj)
  p <- massqc::massqc_pca(safe_log_scale(obj, scale = TRUE), color_by = color_col, order_by = "sample_id") +
    ggtitle(paste0("NEG After Outlier Detection (colored by ", color_col, ")")) +
    theme_light()
  render_report_plot(p, obj)
}
```
:::

::: {.callout-tip title="Figure Legend" collapse="true"}
Outlier-detection PCA shows whether sample-level outliers remain after automatic or manual sample removal.

- **Color**: selected by the group/class rule.
- `r plot_mode_label(objects_by_step$outlier$POS %||% objects_by_step$outlier$NEG)`
- Removed samples are listed in the conclusion above the table.
:::

## Missing Value Imputation

```{r}
#| label: imputation-conclusion
#| results: asis
method <- get_param("missing_value_imputation", "impute_method", "knn")
pos_before <- status_value("outlier", "POS", "missing_fraction")
neg_before <- status_value("outlier", "NEG", "missing_fraction")
pos_after <- status_value("impute", "POS", "missing_fraction")
neg_after <- status_value("impute", "NEG", "missing_fraction")
callout(
  "Missing-value imputation conclusion",
  paste0(
    "Missing values were imputed with ", highlight_numbers(toupper(method)), ". ",
    "Missing-value fraction changed from ",
    highlight_numbers(c(paste0("POS ", fmt_pct(pos_before), " to ", fmt_pct(pos_after)),
                        paste0("NEG ", fmt_pct(neg_before), " to ", fmt_pct(neg_after)))), "."
  )
)
```

```{r imputation-table}
impute_tbl <- status_table |>
  filter(step_id %in% c("outlier", "impute")) |>
  select(mode, step, samples, variables, missing_values, missing_fraction, missing_fraction_delta)
knitr::kable(
  impute_tbl |>
    mutate(missing_fraction = fmt_pct(missing_fraction), missing_fraction_delta = fmt_pct(missing_fraction_delta)),
  caption = "Changes after missing-value imputation."
)
```

::: panel-tabset
### POS

```{r}
#| label: impute-pos-pca
#| fig-height: 6
obj <- objects_by_step$impute$POS
if (!is.null(obj)) {
  color_col <- plot_color_column(obj)
  p <- massqc::massqc_pca(safe_log_scale(obj, scale = TRUE), color_by = color_col, order_by = "sample_id") +
    ggtitle(paste0("POS After Imputation (colored by ", color_col, ")")) +
    theme_light()
  render_report_plot(p, obj)
}
```

### NEG

```{r}
#| label: impute-neg-pca
#| fig-height: 6
obj <- objects_by_step$impute$NEG
if (!is.null(obj)) {
  color_col <- plot_color_column(obj)
  p <- massqc::massqc_pca(safe_log_scale(obj, scale = TRUE), color_by = color_col, order_by = "sample_id") +
    ggtitle(paste0("NEG After Imputation (colored by ", color_col, ")")) +
    theme_light()
  render_report_plot(p, obj)
}
```
:::

::: {.callout-tip title="Figure Legend" collapse="true"}
Imputation PCA checks whether missing-value filling changed the global sample structure.

- **Color**: selected by the group/class rule.
- `r plot_mode_label(objects_by_step$impute$POS %||% objects_by_step$impute$NEG)`
- PCA should remain biologically interpretable after imputation; strong shifts may indicate over-imputation.
:::

## Normalization

```{r}
#| label: normalization-conclusion
#| results: asis
method <- get_param("normalization", "norm_method", "svr")
integ <- get_param("normalization", "integ_method", "qc_mean")
pos_vars <- status_value("norm", "POS", "variables")
neg_vars <- status_value("norm", "NEG", "variables")
pos_mv <- status_value("norm", "POS", "missing_fraction")
neg_mv <- status_value("norm", "NEG", "missing_fraction")
callout(
  "Normalization conclusion",
  paste0(
    "Signal normalization used ", highlight_numbers(toupper(method)),
    "; ", highlight_numbers(integration_label()), ". ",
    "The normalized objects contain ", highlight_numbers(c(paste(fmt_int(pos_vars), "POS features"), paste(fmt_int(neg_vars), "NEG features"))),
    " with missing-value fractions of ",
    highlight_numbers(c(paste0("POS ", fmt_pct(pos_mv)), paste0("NEG ", fmt_pct(neg_mv)))), "."
  )
)
```

```{r normalization-table}
norm_tbl <- status_table |>
  filter(step_id %in% c("impute", "norm")) |>
  select(mode, step, samples, variables, missing_values, missing_fraction)
knitr::kable(
  norm_tbl |>
    mutate(missing_fraction = fmt_pct(missing_fraction)),
  caption = "Changes after normalization."
)
```

::: panel-tabset
### POS

```{r}
#| label: norm-pos-plots
#| fig-height: 6
obj <- objects_by_step$norm$POS
if (!is.null(obj)) {
  color_col <- plot_color_column(obj)
  p <- massqc::massqc_pca(safe_log_scale(obj, scale = TRUE), color_by = color_col, order_by = "sample_id") +
    ggtitle(paste0("POS Normalized PCA (colored by ", color_col, ")")) +
    theme_light()
  render_report_plot(p, obj)
}
```

```{r}
#| label: norm-pos-rsd
#| fig-height: 5
obj <- objects_by_step$norm$POS
if (!is.null(obj)) {
  p <- massqc::massqc_cumulative_rsd_plot(obj, rsd_cutoff = 30, color = "black") +
    ggtitle("POS Normalized Cumulative RSD") +
    theme_light()
  render_report_plot(p, obj)
}
```

### NEG

```{r}
#| label: norm-neg-plots
#| fig-height: 6
obj <- objects_by_step$norm$NEG
if (!is.null(obj)) {
  color_col <- plot_color_column(obj)
  p <- massqc::massqc_pca(safe_log_scale(obj, scale = TRUE), color_by = color_col, order_by = "sample_id") +
    ggtitle(paste0("NEG Normalized PCA (colored by ", color_col, ")")) +
    theme_light()
  render_report_plot(p, obj)
}
```

```{r}
#| label: norm-neg-rsd
#| fig-height: 5
obj <- objects_by_step$norm$NEG
if (!is.null(obj)) {
  p <- massqc::massqc_cumulative_rsd_plot(obj, rsd_cutoff = 30, color = "black") +
    ggtitle("NEG Normalized Cumulative RSD") +
    theme_light()
  render_report_plot(p, obj)
}
```
:::

::: {.callout-tip title="Figure Legend" collapse="true"}
Normalization plots show final PCA structure and cumulative RSD after signal normalization.

- **PCA color**: selected by the group/class rule.
- **Cumulative RSD**: the y-axis is the percentage of features with RSD below each x-axis threshold.
- `r plot_mode_label(objects_by_step$norm$POS %||% objects_by_step$norm$NEG)`
:::

------------------------------------------------------------------------

# Technical Details

::: {.callout-note collapse="true"}
## Mass Dataset Status

This folded section prints the raw `mass_dataset` status for reproducibility and troubleshooting.

::: panel-tabset
### POS

```{r}
#| label: mass-dataset-status-pos
#| results: asis
for (i in seq_len(nrow(step_info))) {
  step_id <- step_info$step_id[[i]]
  obj <- objects_by_step[[step_id]]$POS
  cat("\n\n#### ", step_info$step_label[[i]], "\n\n", sep = "")
  if (is.null(obj)) {
    cat("*Object not available.*\n")
  } else {
    cat("```text\n")
    cat(paste(capture.output(print(obj)), collapse = "\n"))
    cat("\n```\n")
  }
}
```

### NEG

```{r}
#| label: mass-dataset-status-neg
#| results: asis
for (i in seq_len(nrow(step_info))) {
  step_id <- step_info$step_id[[i]]
  obj <- objects_by_step[[step_id]]$NEG
  cat("\n\n#### ", step_info$step_label[[i]], "\n\n", sep = "")
  if (is.null(obj)) {
    cat("*Object not available.*\n")
  } else {
    cat("```text\n")
    cat(paste(capture.output(print(obj)), collapse = "\n"))
    cat("\n```\n")
  }
}
```
:::
:::

------------------------------------------------------------------------

# Citation

Please cite the following papers if you use MetMiner in your research:

> **MetMiner**: Wang, X., Liang, S., Yang, W., Yu, K., Liang, F., Zhao, B., Zhu, X., Zhou, C., Mur, L.A.J., Roberts, J.A., Zhang, J., Zhang, X., 2024.
> MetMiner: A user-friendly pipeline for large-scale plant metabolomics data analysis.
> *J. Integr. Plant Biol.* 66, 2329-2345.
> [doi:10.1111/jipb.13774](https://doi.org/10.1111/jipb.13774)

> **TidyMass2**: Wang, X., Liu, Y., Jiang, C., Huang, Z., Yan, H., Wong, S.H., Johnson, C.H., Zhang, J., Ge, Y., Zhang, F., Zhang, J., Lai, R., Gao, P., Zhang, X., Shen, X., 2026.
> TidyMass2: advancing LC-MS untargeted metabolomics through metabolite origin inference and metabolic feature-based functional module analysis.
> *Nat. Commun.* 17, 1755.
> [doi:10.1038/s41467-026-68464-7](https://doi.org/10.1038/s41467-026-68464-7)

<br> <br>

::: {.column-screen-inset style="background-color: #f8f9fa; padding: 20px; text-align: center; border-top: 1px solid #dee2e6;"}
<small style="color: #6c757d;"> © 2024-2025 **MetMiner** | State Key Laboratory of Crop Stress Adaptation and Improvement<br> Henan University, Kaifeng 475004, China </small>
:::
