---
title: "Differential Analysis and Enrichment Report"
date: last-modified
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    self-contained: true
    code-tools: false
    smooth-scroll: true
params:
  work_dir: null
  metadata: null
  max_tabs: 6
editor:
  markdown:
    wrap: sentence
---

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  fig.width = 10,
  fig.height = 6
)

library(dplyr)
library(plotly)
library(DT)
library(ggplot2)

`%||%` <- function(x, y) if (is.null(x) || length(x) == 0 || all(is.na(x))) y else x
has_text <- function(x) !is.null(x) && length(x) > 0 && !is.na(x[1]) && nzchar(trimws(as.character(x[1])))
has_text_vec <- function(x) {
  if (is.null(x) || length(x) == 0) return(logical())
  !is.na(x) & nzchar(trimws(as.character(x)))
}

work_dir <- params$work_dir %||% "."
tables_dir <- file.path(work_dir, "tables")
status_dir <- file.path(work_dir, "status")
config_dir <- file.path(work_dir, "config")
max_tabs <- params$max_tabs %||% 6

read_tsv <- function(path) {
  if (!file.exists(path)) return(data.frame())
  if (file.info(path)$size <= 3) return(data.frame())
  tryCatch(
    utils::read.delim(path, stringsAsFactors = FALSE, check.names = FALSE, quote = "\"", comment.char = ""),
    error = function(e) data.frame()
  )
}

read_json <- function(path) {
  if (!file.exists(path) || !requireNamespace("jsonlite", quietly = TRUE)) return(NULL)
  jsonlite::fromJSON(path, simplifyVector = FALSE)
}

fmt_int <- function(x) {
  ifelse(is.na(x), NA_character_, format(as.integer(round(x)), big.mark = ","))
}

fmt_num <- function(x, digits = 3) {
  ifelse(is.na(x), NA_character_, signif(as.numeric(x), digits))
}

pick_col <- function(x, candidates) {
  hit <- intersect(candidates, colnames(x))
  if (length(hit) == 0) NA_character_ else hit[1]
}

safe_log10 <- function(x) {
  -log10(pmax(suppressWarnings(as.numeric(x)), .Machine$double.xmin))
}

datatable_report <- function(x, page_length = 10) {
  if (nrow(x) == 0) return(htmltools::p("No rows available."))
  DT::datatable(
    x,
    rownames = FALSE,
    filter = "top",
    extensions = c("Buttons"),
    options = list(
      scrollX = TRUE,
      pageLength = page_length,
      dom = "Bfrtip",
      buttons = c("copy", "csv")
    )
  )
}

highlight <- function(x) {
  paste0("<span style=\"color:#0d6efd;font-weight:700;\">", x, "</span>")
}

make_unique_label <- function(name, id) {
  out <- ifelse(has_text_vec(name), paste0(name, " (", id, ")"), id)
  out
}

volcano_plot <- function(dam_result, fc_cutoff = 1.2, p_cutoff = 0.05, p_column = "p_value") {
  x <- as.data.frame(dam_result %||% data.frame(), stringsAsFactors = FALSE)
  if (nrow(x) == 0) return(plotly::plot_ly())
  x$log2_fc <- suppressWarnings(as.numeric(x$log2_fc %||% log2(x$fc)))
  x$neg_log10_p <- safe_log10(x$p_value)
  x$neg_log10_fdr <- safe_log10(x$p_value_adjust)
  y_col <- if (identical(p_column, "p_value_adjust")) "neg_log10_fdr" else "neg_log10_p"
  label <- if ("Compound.name" %in% colnames(x)) as.character(x$Compound.name) else as.character(x$variable_id)
  label[!has_text_vec(label)] <- as.character(x$variable_id[!has_text_vec(label)])
  x$hover_text <- paste0(
    "Metabolite/feature: ", x$variable_id,
    "<br>Compound: ", label,
    if ("mz" %in% colnames(x)) paste0("<br>m/z: ", signif(suppressWarnings(as.numeric(x$mz)), 6)) else "",
    if ("rt" %in% colnames(x)) paste0("<br>RT: ", signif(suppressWarnings(as.numeric(x$rt)), 5)) else "",
    if ("KEGG.ID" %in% colnames(x)) paste0("<br>KEGG: ", x$KEGG.ID) else "",
    if ("PlantCyc.ID" %in% colnames(x)) paste0("<br>PlantCyc: ", x$PlantCyc.ID) else "",
    if ("Lab.ID" %in% colnames(x)) paste0("<br>Lab ID: ", x$Lab.ID) else "",
    if ("Adduct" %in% colnames(x)) paste0("<br>Adduct: ", x$Adduct) else "",
    "<br>log2FC: ", round(x$log2_fc, 3),
    "<br>p: ", signif(suppressWarnings(as.numeric(x$p_value)), 3),
    "<br>FDR: ", signif(suppressWarnings(as.numeric(x$p_value_adjust)), 3),
    "<br>Change: ", x$change
  )

  p <- ggplot2::ggplot(x, ggplot2::aes(x = log2_fc, y = .data[[y_col]], color = change, key = variable_id, text = hover_text)) +
    ggplot2::geom_point(alpha = 0.75, size = 1.8) +
    ggplot2::geom_vline(xintercept = c(-log2(fc_cutoff), log2(fc_cutoff)), linetype = "dashed", color = "grey55") +
    ggplot2::geom_hline(yintercept = -log10(p_cutoff), linetype = "dashed", color = "grey55") +
    ggplot2::scale_color_manual(values = c("Up" = "#c0392b", "Down" = "#2f6fbb", "Not significant" = "#90979f")) +
    ggplot2::labs(x = "log2 fold change", y = if (identical(p_column, "p_value_adjust")) "-log10(FDR)" else "-log10(p)", color = NULL) +
    ggplot2::theme_minimal(base_size = 12)

  plotly::ggplotly(p, tooltip = "text", source = "dam_volcano") |>
    plotly::layout(legend = list(orientation = "h", x = 0, y = -0.18))
}

plot_img_tag <- function(p, width = 10, height = 6) {
  img <- tempfile(fileext = ".png")
  ggplot2::ggsave(img, plot = p, width = width, height = height, dpi = 150, bg = "white")
  htmltools::tags$img(
    src = knitr::image_uri(img),
    style = "width:100%;max-width:980px;display:block;margin:0 auto 1rem auto;"
  )
}

enrichment_scatter <- function(x, comparison_id, database_label = "KEGG") {
  if (nrow(x) == 0) return(htmltools::p("No enrichment rows available."))
  x$p_value <- suppressWarnings(as.numeric(x$p_value))
  x$p_value_adjust <- suppressWarnings(as.numeric(x$p_value_adjust))
  x$mapped_number <- suppressWarnings(as.numeric(x$mapped_number))
  x$mapped_percentage <- suppressWarnings(as.numeric(x$mapped_percentage))
  x$all_number <- suppressWarnings(as.numeric(x$all_number %||% x$background_number %||% x$mapped_number))
  x <- x |>
    arrange(.data$p_value, .data$p_value_adjust, desc(.data$mapped_number))
  if (requireNamespace("metpath", quietly = TRUE)) {
    metpath_plot <- tryCatch({
      result <- x
      if (!"all_number" %in% colnames(result)) result$all_number <- result$mapped_number
      result$all_number <- suppressWarnings(as.numeric(result$all_number))
      result$all_number[is.na(result$all_number) | result$all_number <= 0] <- result$mapped_number[is.na(result$all_number) | result$all_number <= 0]
      if (!"pathway_name" %in% colnames(result)) result$pathway_name <- result$pathway_id
      if (!"p_value_adjust" %in% colnames(result)) result$p_value_adjust <- result$p_value
      parameter <- tryCatch(
        methods::new(
          Class = "tidymass_parameter",
          pacakge_name = "MetMiner",
          function_name = "metminer_differential_enrichment_report()",
          parameter = list(
            comparison_id = comparison_id,
            database = database_label,
            y_axis = "p_value",
            cutoff = 0.05
          ),
          time = Sys.time()
        ),
        error = function(e) NULL
      )
      if (is.null(parameter)) stop("Could not create tidymass parameter.", call. = FALSE)
      enrich_obj <- methods::new(
        Class = "enrich_result",
        pathway_database = database_label,
        pathway_version = "",
        result = result,
        parameter = parameter
      )
      p <- metpath::enrich_scatter_plot(
        enrich_obj,
        x_axis = "mapped_percentage",
        y_axis = "p_value",
        point_size = "all_number",
        x_axis_cutoff = 0,
        y_axis_cutoff = 0.05,
        label = TRUE,
        label_size = 3.5
      )
      if (is.null(p)) stop("No pathways are enriched.", call. = FALSE)
      p <- p +
        ggplot2::labs(title = NULL) +
        ggplot2::theme(
          plot.title = ggplot2::element_blank(),
          legend.position = "right"
        )
      plot_img_tag(p)
    }, error = function(e) NULL)
    if (!is.null(metpath_plot)) return(metpath_plot)
  }
  x$pathway_label <- make_unique_label(x$pathway_name, x$pathway_id)
  x$neg_log10_p <- safe_log10(x$p_value)
  x$is_significant <- x$p_value < 0.05 & x$mapped_percentage > 0
  p <- ggplot2::ggplot(
    x,
    ggplot2::aes(
      x = mapped_percentage,
      y = neg_log10_p,
      size = all_number,
      fill = is_significant
    )
  ) +
    ggplot2::geom_hline(yintercept = -log10(0.05), color = "grey70", linetype = 2) +
    ggplot2::geom_vline(xintercept = 0, color = "grey70", linetype = 2) +
    ggplot2::geom_point(shape = 21, color = "black", alpha = 0.9) +
    ggplot2::scale_fill_manual(values = c("TRUE" = "#EE0000FF", "FALSE" = "grey70"), guide = "none") +
    ggplot2::labs(x = "Mapped percentage (%)", y = "-log(P-values, 10)", size = "Pathway size") +
    ggplot2::theme_bw(base_size = 12) +
    ggplot2::theme(panel.grid = ggplot2::element_blank())
  if (requireNamespace("ggrepel", quietly = TRUE)) {
    p <- p + ggrepel::geom_text_repel(
      data = x[x$is_significant, , drop = FALSE],
      ggplot2::aes(label = pathway_label),
      size = 3.5,
      show.legend = FALSE
    )
  }
  plot_img_tag(p)
}

filter_enrichment_p <- function(x, p_cutoff = 0.05) {
  x <- as.data.frame(x %||% data.frame(), stringsAsFactors = FALSE)
  if (nrow(x) == 0 || !"p_value" %in% colnames(x)) return(x)
  x$p_value <- suppressWarnings(as.numeric(x$p_value))
  x[x$p_value < p_cutoff, , drop = FALSE]
}

drop_description_cols <- function(x) {
  x <- as.data.frame(x %||% data.frame(), stringsAsFactors = FALSE)
  x[, setdiff(colnames(x), c("describtion", "description", "Description")), drop = FALSE]
}

clean_enrichment_table <- function(x) {
  x <- drop_description_cols(x)
  x[, setdiff(colnames(x), "query_id"), drop = FALSE]
}

metric_value <- function(x, metric, default = NA_integer_) {
  if (nrow(x) == 0 || !all(c("metric", "value") %in% colnames(x))) return(default)
  hit <- match(metric, as.character(x$metric))
  if (is.na(hit)) default else suppressWarnings(as.integer(x$value[hit]))
}

diff_table <- read_tsv(file.path(tables_dir, "10.differential_result.tsv"))
diff_summary <- read_tsv(file.path(tables_dir, "10.differential_summary.tsv"))
enrich_table <- read_tsv(file.path(tables_dir, "09.enrichment_result.tsv"))
enrich_p <- read_tsv(file.path(tables_dir, "09.enrichment_p_0_05.tsv"))
enrich_fdr <- read_tsv(file.path(tables_dir, "09.enrichment_fdr_0_05.tsv"))
query_background <- read_tsv(file.path(tables_dir, "09.enrichment_query_background.tsv"))
plantcyc_enrich_table <- read_tsv(file.path(tables_dir, "09.plantcyc_enrichment_result.tsv"))
plantcyc_enrich_p <- read_tsv(file.path(tables_dir, "09.plantcyc_enrichment_p_0_05.tsv"))
plantcyc_enrich_fdr <- read_tsv(file.path(tables_dir, "09.plantcyc_enrichment_fdr_0_05.tsv"))
plantcyc_query_background <- read_tsv(file.path(tables_dir, "09.plantcyc_enrichment_query_background.tsv"))
clean_matrix <- read_tsv(file.path(tables_dir, "08.clean_annotation_matrix.tsv"))
sample_info <- read_tsv(file.path(config_dir, "sample_info.tsv"))
clean_annotation_summary <- read_tsv(file.path(tables_dir, "08.clean_annotation_dataset_summary.tsv"))
annotation_filtering_summary <- read_tsv(file.path(tables_dir, "08.annotation_filtering_summary.tsv"))
redundancy_audit <- read_tsv(file.path(tables_dir, "08.redundancy_audit.tsv"))
compare_group <- read_tsv(file.path(config_dir, "compare_group.tsv"))
diff_status <- read_json(file.path(status_dir, "differential_analysis.json"))
enrich_status <- read_json(file.path(status_dir, "enrichment.json"))

preferred_comparisons <- c(
  "OEL_7D_vs_EV_7D",
  "OEH_7D_vs_EV_7D",
  "OEH_7D_vs_OEL_7D",
  "OEL_12D_vs_EV_12D",
  "OEH_12D_vs_EV_12D",
  "OEH_12D_vs_OEL_12D"
)

available_comparisons <- if (nrow(compare_group) > 0 && "comparison_id" %in% colnames(compare_group)) {
  unique(as.character(compare_group$comparison_id))
} else if (nrow(diff_summary) > 0 && "comparison_id" %in% colnames(diff_summary)) {
  unique(as.character(diff_summary$comparison_id))
} else {
  unique(as.character(diff_table$comparison_id %||% enrich_table$comparison_id %||% character()))
}
available_comparisons <- available_comparisons[has_text_vec(available_comparisons)]
comparison_order <- c(
  preferred_comparisons[preferred_comparisons %in% available_comparisons],
  setdiff(available_comparisons, preferred_comparisons)
)
display_comparisons <- head(comparison_order, max_tabs)

dam_params <- diff_status$parameters %||% list()
fc_cutoff <- as.numeric(dam_params$default_fc_cutoff %||% 1.2)
p_cutoff <- as.numeric(dam_params$default_p_cutoff %||% 0.05)
p_col <- "p_value"

clean_records <- metric_value(clean_annotation_summary, "clean_metabolites", metric_value(annotation_filtering_summary, "final_records", nrow(clean_matrix)))
audit_records <- metric_value(annotation_filtering_summary, "audit_records", nrow(redundancy_audit))
dropped_records <- metric_value(annotation_filtering_summary, "dropped_records", if (nrow(redundancy_audit) > 0 && "keep" %in% colnames(redundancy_audit)) sum(!redundancy_audit$keep, na.rm = TRUE) else NA_integer_)
cross_polarity_dropped <- if (nrow(redundancy_audit) > 0 && all(c("keep", "cross_polarity_group") %in% colnames(redundancy_audit))) {
  sum(!redundancy_audit$keep & has_text_vec(redundancy_audit$cross_polarity_group), na.rm = TRUE)
} else {
  NA_integer_
}
recurrent_review_records <- if (nrow(redundancy_audit) > 0 && "recurrent_status" %in% colnames(redundancy_audit)) {
  sum(has_text_vec(redundancy_audit$recurrent_status) & redundancy_audit$recurrent_status != "none", na.rm = TRUE)
} else {
  NA_integer_
}
```

## Summary

This report summarizes differential abundance metabolite analysis and KEGG enrichment for the clean, non-redundant annotation result.

The differential analysis is run on the final clean metabolite/sub-network table rather than on every raw feature.
This keeps the statistical table aligned with the annotation review result: one row represents a final metabolite-like unit, with feature-network pseudo area used where available.

The DAM cutoff used in this report is **`r htmltools::HTML(highlight(paste0("FC >= ", fc_cutoff)))`** and **`r htmltools::HTML(highlight(paste0("P < ", p_cutoff)))`**.

For readability and browser performance, the detailed DAM result section renders interactive tabs for the first `r length(display_comparisons)` comparison groups, with a maximum of `r max_tabs`.
The preferred display order is `OEL_7D_vs_EV_7D`, `OEH_7D_vs_EV_7D`, `OEH_7D_vs_OEL_7D`, `OEL_12D_vs_EV_12D`, `OEH_12D_vs_EV_12D`, and `OEH_12D_vs_OEL_12D`.
If fewer than six groups are available, all available groups are shown.
All comparison groups and all rows are still available in the TSV outputs under the `tables/` directory.

::: {.callout-important}
## Clean Annotation Result

The final non-redundant annotation table contains **`r fmt_int(clean_records)`** records.
The redundancy audit tracked **`r fmt_int(audit_records)`** candidate records, including **`r fmt_int(dropped_records)`** dropped records and **`r fmt_int(cross_polarity_dropped)`** cross-polarity redundant records.
Feature-network recurrent context flagged **`r fmt_int(recurrent_review_records)`** records for review.
:::

```{r}
#| label: overview-values

summary_display <- diff_summary
if (nrow(summary_display) > 0 && "comparison_id" %in% colnames(summary_display)) {
  summary_display$.display_order <- match(as.character(summary_display$comparison_id), comparison_order)
  summary_display <- summary_display[order(summary_display$.display_order, na.last = TRUE), , drop = FALSE]
}
summary_keep <- intersect(
  c("comparison_id", "description", "case_group", "control_group", "significant", "up", "down"),
  colnames(summary_display)
)
summary_display <- summary_display[, summary_keep, drop = FALSE]
datatable_report(summary_display, page_length = 18)
```

::: {.callout-warning}
## KEGG Enrichment Background

For species without a dedicated KEGG organism pathway database, this workflow uses all KEGG IDs detected in the clean annotation result as the enrichment background.
This is useful for exploratory screening, but it is less conservative than a species-specific detectable pathway universe and may increase false-positive pathway signals.
The report therefore keeps the enrichment display focused on the exploratory `P < 0.05` result tables.
:::

## DAM Results

```{r}
#| label: comparison-tabs
#| results: markup

comparison_panel <- function(comparison_id, active = FALSE) {
  dam_i <- diff_table[as.character(diff_table$comparison_id) == comparison_id, , drop = FALSE]
  enrich_i <- enrich_table[as.character(enrich_table$comparison_id) == comparison_id, , drop = FALSE]
  plantcyc_i <- plantcyc_enrich_table[as.character(plantcyc_enrich_table$comparison_id) == comparison_id, , drop = FALSE]
  summary_i <- diff_summary[as.character(diff_summary$comparison_id) == comparison_id, , drop = FALSE]
  query_i <- query_background[as.character(query_background$comparison_id) == comparison_id, , drop = FALSE]
  plantcyc_query_i <- plantcyc_query_background[as.character(plantcyc_query_background$comparison_id) == comparison_id, , drop = FALSE]

  tab_id <- paste0("dam-tab-", gsub("[^A-Za-z0-9_-]+", "-", comparison_id))
  summary_text <- if (nrow(summary_i) > 0) {
    paste0(
      "DAMs: <strong>", fmt_int(summary_i$significant[1]), "</strong> ",
      "(Up: <strong>", fmt_int(summary_i$up[1]), "</strong>; Down: <strong>", fmt_int(summary_i$down[1]), "</strong>). ",
      "Features/metabolites tested: <strong>", fmt_int(summary_i$features_tested[1]), "</strong>."
    )
  } else {
    "DAM summary is not available."
  }

  dam_table_i <- dam_i[dam_i$change %in% c("Up", "Down"), , drop = FALSE]
  dam_keep <- intersect(
    c("comparison_id", "variable_id", "representative_feature", "Compound.name", "KEGG.ID", "PlantCyc.ID",
      "fc", "log2_fc", "p_value", "p_value_adjust", "change", "mz", "rt", "Super_class", "Class", "Sub_class"),
    colnames(dam_table_i)
  )

  enrich_i$database <- "KEGG"
  plantcyc_i$database <- "PlantCyc"
  enrich_i_p <- filter_enrichment_p(enrich_i, p_cutoff)
  plantcyc_i_p <- filter_enrichment_p(plantcyc_i, p_cutoff)
  enrich_all_i <- dplyr::bind_rows(enrich_i_p, plantcyc_i_p)
  enrich_all_i <- clean_enrichment_table(enrich_all_i)
  enrich_keep <- intersect(
    c("database", "comparison_id", "pathway_id", "pathway_name", "p_value", "p_value_adjust", "background_number",
      "query_number", "mapped_number", "mapped_percentage", "mapped_id"),
    colnames(enrich_all_i)
  )

  htmltools::tags$div(
    id = tab_id,
    class = paste("tab-pane fade", if (active) "show active" else ""),
    role = "tabpanel",
    `aria-labelledby` = paste0(tab_id, "-tab"),
    htmltools::tags$div(
      class = "pt-3",
      htmltools::tags$p(htmltools::HTML(summary_text)),
      htmltools::tags$h4("Interactive Volcano Plot"),
      volcano_plot(dam_i, fc_cutoff = fc_cutoff, p_cutoff = p_cutoff, p_column = p_col),
      htmltools::tags$h4("KEGG Enrichment Scatter Plot"),
      enrichment_scatter(enrich_i, comparison_id, database_label = "KEGG"),
      htmltools::tags$h4("PlantCyc Enrichment Result"),
      enrichment_scatter(plantcyc_i, comparison_id, database_label = "PlantCyc"),
      htmltools::tags$h4("Differential Analysis Table"),
      datatable_report(dam_table_i[, dam_keep, drop = FALSE], page_length = 10),
      htmltools::tags$h4("Enrichment Table"),
      datatable_report(enrich_all_i[, enrich_keep, drop = FALSE], page_length = 10)
    )
  )
}

if (length(display_comparisons) == 0) {
  htmltools::p("No comparison result is available.")
} else {
  nav_items <- lapply(seq_along(display_comparisons), function(i) {
    comparison_id <- display_comparisons[[i]]
    tab_id <- paste0("dam-tab-", gsub("[^A-Za-z0-9_-]+", "-", comparison_id))
    htmltools::tags$li(
      class = "nav-item",
      role = "presentation",
      htmltools::tags$button(
        class = paste("nav-link", if (i == 1) "active" else ""),
        id = paste0(tab_id, "-tab"),
        `data-bs-toggle` = "tab",
        `data-bs-target` = paste0("#", tab_id),
        type = "button",
        role = "tab",
        `aria-controls` = tab_id,
        `aria-selected` = if (i == 1) "true" else "false",
        comparison_id
      )
    )
  })

  panels <- lapply(seq_along(display_comparisons), function(i) {
    comparison_panel(display_comparisons[[i]], active = i == 1)
  })

  htmltools::tagList(
    htmltools::tags$div(
      class = "panel-tabset metminer-dam-tabset",
      htmltools::tags$ul(class = "nav nav-tabs", role = "tablist", nav_items),
      htmltools::tags$div(class = "tab-content", panels)
    )
  )
}
```

## Threshold Tables

### KEGG Enrichment P < 0.05

```{r}
#| label: enrichment-p-table

datatable_report(clean_enrichment_table(enrich_p), page_length = 10)
```

### PlantCyc Enrichment P < 0.05

```{r}
#| label: plantcyc-enrichment-p-table

datatable_report(clean_enrichment_table(plantcyc_enrich_p), page_length = 10)
```

## Output Files

```{r}
#| label: output-files

files <- data.frame(
  file = c(
    "tables/10.differential_result.tsv",
    "tables/10.differential_summary.tsv",
    "tables/09.enrichment_result.tsv",
    "tables/09.enrichment_p_0_05.tsv",
    "tables/09.enrichment_query_background.tsv",
    "tables/09.plantcyc_enrichment_result.tsv",
    "tables/09.plantcyc_enrichment_p_0_05.tsv",
    "tables/09.plantcyc_enrichment_query_background.tsv"
  ),
  description = c(
    "Full differential abundance result table for all comparison groups.",
    "One-row summary per comparison group.",
    "Full KEGG enrichment table.",
    "KEGG enrichment table filtered at P < 0.05.",
    "Query/background diagnostic table for KEGG enrichment.",
    "Full PlantCyc enrichment table.",
    "PlantCyc enrichment table filtered at P < 0.05.",
    "Query/background diagnostic table for PlantCyc enrichment."
  )
)
datatable_report(files, page_length = 10)
```
