---
title: "Feature Annotation Report"
date: last-modified
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    self-contained: true
    code-tools: false
    smooth-scroll: true
params:
  work_dir: null
  metadata: null
editor:
  markdown:
    wrap: sentence
---

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  fig.width = 10,
  fig.height = 6
)

library(ggplot2)
library(dplyr)
library(tidyr)
library(plotly)
library(massdataset)

`%||%` <- function(x, y) if (is.null(x) || length(x) == 0 || all(is.na(x))) y else x
has_text <- function(x) !is.null(x) && length(x) > 0 && !is.na(x[1]) && nzchar(trimws(as.character(x[1])))

work_dir <- params$work_dir %||% "."
tables_dir <- file.path(work_dir, "tables")
status_dir <- file.path(work_dir, "status")

read_tsv <- function(name) {
  path <- file.path(tables_dir, name)
  if (!file.exists(path)) return(data.frame())
  utils::read.delim(path, stringsAsFactors = FALSE, check.names = FALSE, quote = "\"", comment.char = "")
}

read_json <- function(name) {
  path <- if (grepl("/", name, fixed = TRUE)) file.path(work_dir, name) else file.path(status_dir, name)
  if (!file.exists(path) || !requireNamespace("jsonlite", quietly = TRUE)) return(NULL)
  jsonlite::fromJSON(path, simplifyVector = FALSE)
}

fmt_int <- function(x) {
  ifelse(is.na(x), NA_character_, format(as.integer(round(x)), big.mark = ","))
}

fmt_pct <- function(x) {
  ifelse(is.na(x), NA_character_, sprintf("%.1f%%", as.numeric(x)))
}

metric_value <- function(tab, metric, default = NA_real_) {
  if (nrow(tab) == 0 || !"metric" %in% colnames(tab) || !"value" %in% colnames(tab)) return(default)
  hit <- which(tab$metric == metric)
  if (length(hit) == 0) return(default)
  suppressWarnings(as.numeric(tab$value[hit[1]]))
}

extract_ids <- function(tab, candidates) {
  if (nrow(tab) == 0) return(character())
  col <- intersect(candidates, colnames(tab))[1]
  if (!has_text(col)) return(character())
  ids <- unlist(strsplit(paste(as.character(tab[[col]]), collapse = ";"), ";|\\|\\||,|\\s+", perl = TRUE), use.names = FALSE)
  ids <- trimws(ids)
  ids <- ids[!is.na(ids) & nzchar(ids)]
  unique(ids)
}

plotly_or_static <- function(p, n = 0) {
  if (is.finite(n) && n <= 1000) {
    plotly::ggplotly(p) |>
      plotly::layout(legend = list(orientation = "v"))
  } else {
    p
  }
}

collapse_text <- function(...) {
  vals <- unlist(list(...), use.names = FALSE)
  vals <- trimws(as.character(vals %||% ""))
  vals <- vals[has_text(vals)]
  vals <- vals[!duplicated(vals)]
  paste(vals, collapse = " | ")
}

top_n_classes <- function(summary_table, dimension, n = 12) {
  x <- summary_table |>
    filter(.data$dimension == dimension) |>
    mutate(
      class_name = ifelse(
        !has_text(.data$class_name) | .data$class_name %in% c("Unclassified", "Other"),
        "Unclassified / Other",
        .data$class_name
      )
    ) |>
    group_by(.data$dimension, .data$class_name) |>
    summarise(
      count = sum(.data$count, na.rm = TRUE),
      percent = sum(.data$percent, na.rm = TRUE),
      metabolite_ids = paste(.data$metabolite_ids[has_text(.data$metabolite_ids)], collapse = ";"),
      compound_names = paste(.data$compound_names[has_text(.data$compound_names)], collapse = ";"),
      .groups = "drop"
    ) |>
    arrange(desc(.data$count), .data$class_name)
  if (nrow(x) == 0) return(x)
  if (nrow(x) > n) {
    classified <- x[x$class_name != "Unclassified / Other", , drop = FALSE]
    unclassified <- x[x$class_name == "Unclassified / Other", , drop = FALSE]
    kept <- classified[seq_len(min(n, nrow(classified))), , drop = FALSE]
    other <- classified[-seq_len(min(n, nrow(classified))), , drop = FALSE]
    if (nrow(other) > 0) {
      other_row <- data.frame(
        dimension = dimension,
        class_name = "Unclassified / Other",
        count = sum(c(other$count, unclassified$count), na.rm = TRUE),
        percent = sum(c(other$percent, unclassified$percent), na.rm = TRUE),
        metabolite_ids = paste(c(other$metabolite_ids, unclassified$metabolite_ids)[has_text(c(other$metabolite_ids, unclassified$metabolite_ids))], collapse = ";"),
        compound_names = paste(utils::head(unlist(strsplit(paste(c(other$compound_names, unclassified$compound_names), collapse = ";"), ";", fixed = TRUE)), 50), collapse = ";"),
        stringsAsFactors = FALSE
      )
      x <- rbind(kept, other_row)
    } else {
      x <- rbind(kept, unclassified)
    }
  }
  x
}

pie_plot <- function(summary_table, dimension) {
  x <- top_n_classes(summary_table, dimension, 12)
  if (nrow(x) == 0) {
    return(plotly::plot_ly())
  }
  plotly::plot_ly(
    data = x,
    labels = ~class_name,
    values = ~count,
    type = "pie",
    text = ~paste0(class_name, " (", count, ", ", sprintf("%.1f", percent), "%)"),
    hovertext = ~paste0(
      class_name,
      "<br>Count: ", count,
      "<br>Percent: ", sprintf("%.2f", percent), "%",
      "<br>Examples: ", compound_names
    ),
    hoverinfo = "text",
    textinfo = "percent",
    sort = FALSE
  ) |>
    plotly::layout(
      title = list(text = paste("ClassyFire", dimension)),
      legend = list(orientation = "v"),
      margin = list(l = 10, r = 10, t = 60, b = 10)
    )
}

classification_detail_table <- function(sort_by = "Super_class") {
  if (nrow(final_annotations) == 0) return(data.frame())
  x <- final_annotations
  for (col in c("metabolite_id", "compound_name", "Super_class", "Class", "Sub_class", "direct_parent")) {
    if (!col %in% colnames(x)) x[[col]] <- NA_character_
  }
  parent_cols <- intersect(
    c("direct_parent", "parent_level_1", "parent_level_2", "parent_level_3", "parent_level_4", "parent_level_5"),
    colnames(x)
  )
  parent_levels <- if (length(parent_cols) > 0) {
    apply(x[, parent_cols, drop = FALSE], 1, function(row) collapse_text(row))
  } else {
    rep("", nrow(x))
  }
  out <- data.frame(
    `metabolite id` = as.character(x$metabolite_id),
    `compound name` = as.character(x$compound_name),
    `super class` = as.character(x$Super_class),
    class = as.character(x$Class),
    subclass = as.character(x$Sub_class),
    `parent levels` = parent_levels,
    check.names = FALSE,
    stringsAsFactors = FALSE
  )
  for (col in colnames(out)) {
    out[[col]][!has_text(out[[col]])] <- ""
  }
  informative <- !(
    out[["super class"]] %in% c("", "Unclassified", "Other") &
      out[["class"]] %in% c("", "Unclassified", "Other") &
      out[["subclass"]] %in% c("", "Unclassified", "Other")
  )
  out <- out[informative, , drop = FALSE]
  sort_col <- switch(sort_by,
    Super_class = "super class",
    Class = "class",
    Sub_class = "subclass",
    "super class"
  )
  out <- out[order(out[[sort_col]], out[["compound name"]], out[["metabolite id"]]), , drop = FALSE]
  rownames(out) <- NULL
  out
}

load_rda_first <- function(path) {
  if (!file.exists(path)) return(NULL)
  e <- new.env(parent = emptyenv())
  nm <- load(path, envir = e)
  if (length(nm) == 0) return(NULL)
  e[[nm[[1]]]]
}

feature_network_from_object <- function(path) {
  obj <- load_rda_first(path)
  if (is.null(obj)) return(data.frame())
  if (is.data.frame(obj)) return(obj)
  net <- tryCatch(obj@other_files$feature_network, error = function(e) NULL)
  if (is.null(net)) return(data.frame())
  as.data.frame(net, stringsAsFactors = FALSE)
}

network_display_data <- function(edges, max_edges = 700) {
  edges <- as.data.frame(edges %||% data.frame(), stringsAsFactors = FALSE)
  if (nrow(edges) == 0 || !all(c("from", "to") %in% colnames(edges))) {
    return(list(nodes = data.frame(), edges = data.frame()))
  }
  if (!"type" %in% colnames(edges)) edges$type <- "Feature relation"
  if (!"relation_score" %in% colnames(edges)) edges$relation_score <- NA_real_
  edges$relation_score <- suppressWarnings(as.numeric(edges$relation_score))
  edges <- edges[has_text(edges$from) & has_text(edges$to), , drop = FALSE]
  edges <- edges[order(-ifelse(is.finite(edges$relation_score), edges$relation_score, 0)), , drop = FALSE]
  if (nrow(edges) > max_edges) edges <- edges[seq_len(max_edges), , drop = FALSE]

  ids <- unique(c(edges$from, edges$to))
  degree <- sort(table(c(edges$from, edges$to)), decreasing = TRUE)
  nodes <- data.frame(
    id = ids,
    label = ids,
    value = as.numeric(degree[match(ids, names(degree))]),
    group = ifelse(grepl("^pos::", ids), "Positive feature",
      ifelse(grepl("^neg::", ids), "Negative feature", "Feature")
    ),
    title = paste0(ids, "<br>Degree: ", as.numeric(degree[match(ids, names(degree))])),
    stringsAsFactors = FALSE
  )
  edge_group <- as.character(edges$type)
  edge_color <- dplyr::case_when(
    grepl("isotope", edge_group, ignore.case = TRUE) ~ "#1b9e77",
    grepl("adduct", edge_group, ignore.case = TRUE) ~ "#7570b3",
    grepl("isf|fragment", edge_group, ignore.case = TRUE) ~ "#d95f02",
    grepl("cross", edge_group, ignore.case = TRUE) ~ "#4d4d4d",
    TRUE ~ "#7f8c8d"
  )
  display_edges <- data.frame(
    from = edges$from,
    to = edges$to,
    label = edge_group,
    title = paste0(
      edge_group,
      "<br>", edges$annotation %||% "",
      "<br>Score: ", sprintf("%.3f", edges$relation_score)
    ),
    arrows = "to",
    color = edge_color,
    stringsAsFactors = FALSE
  )
  list(nodes = nodes, edges = display_edges)
}

feature_network_widget <- function(edges, title = "Feature network", max_edges = 700) {
  display <- network_display_data(edges, max_edges = max_edges)
  if (nrow(display$nodes) == 0 || nrow(display$edges) == 0) {
    return(plotly::plot_ly() |> plotly::layout(title = list(text = paste(title, "not available"))))
  }
  if (requireNamespace("visNetwork", quietly = TRUE)) {
    return(
      visNetwork::visNetwork(display$nodes, display$edges, height = "680px", main = title) |>
        visNetwork::visNodes(shape = "dot", font = list(size = 15)) |>
        visNetwork::visGroups(groupname = "Positive feature", color = "#2c7fb8") |>
        visNetwork::visGroups(groupname = "Negative feature", color = "#7a5195") |>
        visNetwork::visGroups(groupname = "Feature", color = "#7f8c8d") |>
        visNetwork::visEdges(smooth = FALSE, font = list(align = "middle", size = 10)) |>
        visNetwork::visOptions(highlightNearest = TRUE, nodesIdSelection = TRUE) |>
        visNetwork::visInteraction(dragNodes = TRUE, dragView = TRUE, hover = TRUE, tooltipDelay = 80) |>
        visNetwork::visPhysics(enabled = TRUE, stabilization = TRUE) |>
        visNetwork::visLegend(useGroups = TRUE, position = "right")
    )
  }
  graph <- network_display_data(edges, max_edges = min(max_edges, 200))
  p <- ggplot(graph$edges, aes(from, to, color = label)) +
    geom_count(show.legend = TRUE) +
    labs(title = title, x = "From", y = "To") +
    theme_light() +
    theme(axis.text.x = element_blank(), axis.text.y = element_blank())
  plotly::ggplotly(p)
}

ms2_summary <- read_tsv("05.ms2_attachment_summary.tsv")
network_summary <- read_tsv("06.feature_network_summary.tsv")
annotation_summary <- read_tsv("07.annotation_summary.tsv")
filter_summary <- read_tsv("08.annotation_filtering_summary.tsv")
classification_status <- read_tsv("09.classification_status.tsv")
classification_summary <- read_tsv("09.classification_summary.tsv")
classification_records <- read_tsv("09.classification_records.tsv")
final_annotations <- read_tsv("08.final_annotations.tsv")
collapse_annotations <- read_tsv("08.collapse_annotations.tsv")
expand_annotations <- read_tsv("08.expand_annotations.tsv")
redundancy_audit <- read_tsv("08.redundancy_audit.tsv")
feature_network_pos <- feature_network_from_object(file.path(work_dir, "objects", "06.object_pos_feature_network.rda"))
feature_network_neg <- feature_network_from_object(file.path(work_dir, "objects", "06.object_neg_feature_network.rda"))
feature_network_merged <- feature_network_from_object(file.path(work_dir, "objects", "06.merged_feature_network.rda"))

annotation_status <- read_json("annotation.json")
filter_status <- read_json("annotation_filtering.json")
classification_json <- read_json("classification.json")
resource_plan <- read_json("config/annotation_resources.json")

used_databases <- as.character(unlist(annotation_status$databases %||% character(), use.names = FALSE))
resource_species <- resource_plan$species %||% list()
kegg_decision <- resource_plan$resources$kegg$decision %||% NA_character_
has_species_kegg <- kegg_decision %in% c("existing_database", "build_kegg_organism", "built_in_database")
kegg_background_ids <- extract_ids(collapse_annotations, c("KEGG.ID", "KEGG_ID", "kegg_id"))
if (length(kegg_background_ids) == 0) {
  kegg_background_ids <- extract_ids(expand_annotations, c("KEGG.ID", "KEGG_ID", "kegg_id"))
}
if (length(kegg_background_ids) == 0) {
  kegg_background_ids <- extract_ids(final_annotations, c("KEGG.ID", "KEGG_ID", "kegg_id"))
}
use_detected_kegg_warning <- !isTRUE(has_species_kegg) && length(kegg_background_ids) > 0
resource_table <- data.frame(
  resource = c("PlantCyc", "KEGG", "Public MS2"),
  decision = c(
    resource_plan$resources$plantcyc$decision %||% NA_character_,
    resource_plan$resources$kegg$decision %||% NA_character_,
    paste(as.character(unlist(resource_plan$resources$public_ms2 %||% character(), use.names = FALSE)), collapse = "; ")
  ),
  reason = c(
    resource_plan$resources$plantcyc$reason %||% NA_character_,
    resource_plan$resources$kegg$reason %||% NA_character_,
    "Public experimental MS/MS libraries are used as broad spectral evidence together with plant knowledge-base candidates."
  ),
  stringsAsFactors = FALSE
)

final_records <- metric_value(filter_summary, "final_records", nrow(final_annotations))
audit_records <- metric_value(filter_summary, "audit_records", nrow(redundancy_audit))
dropped_records <- metric_value(filter_summary, "dropped_records")
cross_removed <- metric_value(filter_summary, "cross_polarity_removed")
recurrent_flagged <- metric_value(filter_summary, "recurrent_flagged")
```

::: {.callout-note icon="false" appearance="simple"}
## Report Information

**Software**: [MetMiner by Wang et al., 2024](https://github.com/ShawnWx2019/MetMiner), powered by [TidyMass2 by Wang et al., 2026](https://www.nature.com/articles/s41467-026-68464-7)

**Date**: `r format(Sys.Date(), '%Y-%m-%d')`

**Generated By**: MetMiner Snakemake pipeline
:::

------------------------------------------------------------------------

# Introduction

This report summarizes the feature annotation workflow after normalization and feature-network construction.
It covers MS2 attachment, feature-network evidence, database annotation, non-redundant annotation filtering, and ClassyFire-based classification of the final clean annotation table.

## Annotation Database Strategy

For this project, the reported species is **`r resource_species$common_name %||% "not specified"`** (`r resource_species$latin_name %||% "Latin name not specified"`).
MetMiner used the following annotation databases:

```{r used-databases}
database_table <- data.frame(
  database = used_databases,
  stringsAsFactors = FALSE
)
knitr::kable(database_table, caption = "Annotation databases used in this run.")
```

```{r resource-plan}
knitr::kable(resource_table, caption = "Species-aware annotation resource plan.")
```

MetMiner recommends this layered strategy because plant metabolomics annotation has two different evidence needs.
First, PlantCyc or species-specific plant databases provide biologically plausible plant metabolites and pathway-aware compound candidates, which helps avoid treating every public-library spectral hit as equally relevant to the organism.
Second, HMDB, MoNA, and MassBank add experimental MS/MS evidence and broaden coverage when the study species does not yet have a dedicated PlantCyc or KEGG resource.
The feature-network and redundancy-filtering steps then reduce adduct, isotope, in-source fragment, recurrent-ion, and cross-polarity duplication before the final clean annotation table is used for classification, differential analysis, and enrichment.

```{r kegg-fallback-warning, results='asis'}
if (isTRUE(use_detected_kegg_warning)) {
  cat("::: {.callout-warning appearance=\"simple\" icon=\"true\"}\\n")
  cat("## Exploratory KEGG Enrichment Background Warning\\n\\n")
  cat("A species-specific KEGG organism database was not available for this run. ")
  cat("The preferred KEGG enrichment background is the species-specific detectable pathway metabolite universe, built from the organism's KEGG pathways and the compounds that can be annotated under the same LC-MS workflow. ")
  cat("For this project, MetMiner can only use the **", length(kegg_background_ids), "** KEGG IDs detected in the clean annotation results as a fallback background. ")
  cat("This detected-ID background is useful for exploratory screening, but it is conditioned on the observed data and annotation coverage, so pathway p-values may have an elevated false-positive rate and should not be interpreted as confirmatory species-level pathway enrichment.\\n")
  cat(":::\\n")
}
```

------------------------------------------------------------------------

# Annotation Overview

::: {.callout-important appearance="simple" icon="false"}
## Clean Annotation Result

The final non-redundant annotation table contains **`r fmt_int(final_records)`** records.
The redundancy audit tracked **`r fmt_int(audit_records)`** candidate records, including **`r fmt_int(dropped_records)`** dropped records and **`r fmt_int(cross_removed)`** cross-polarity redundant records.
Feature-network recurrent context flagged **`r fmt_int(recurrent_flagged)`** records for review.
:::

```{r annotation-overview-table}
overview <- data.frame(
  metric = c(
    "Final non-redundant records",
    "Redundancy audit records",
    "Expand review rows",
    "Collapse review rows",
    "Dropped redundant records",
    "Cross-polarity removed",
    "Recurrent context flagged"
  ),
  value = c(
    final_records,
    audit_records,
    nrow(expand_annotations),
    nrow(collapse_annotations),
    dropped_records,
    cross_removed,
    recurrent_flagged
  )
)
knitr::kable(overview, caption = "Feature annotation and filtering overview.")
```

```{r annotation-overview-plot}
overview_plot_data <- overview |>
  filter(!is.na(value)) |>
  mutate(metric = factor(metric, levels = rev(metric)))

p <- ggplot(overview_plot_data, aes(value, metric, fill = metric)) +
  geom_col(width = 0.65, show.legend = FALSE) +
  geom_text(aes(label = fmt_int(value)), hjust = -0.08, size = 3.5) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(x = "Count", y = NULL, title = "Annotation Filtering Summary") +
  theme_light()
plotly_or_static(p, max(overview_plot_data$value, na.rm = TRUE))
```

::: {.callout-tip title="Figure Legend" collapse="true"}
The bar plot shows the main outputs of annotation filtering.
Interactive rendering is used when the displayed record count is at or below 1,000; otherwise a static plot is shown.
:::

------------------------------------------------------------------------

# Step-by-Step Feature Annotation

## MS2 Attachment

The pipeline searches for MGF files generated by the raw conversion step and attaches MS2 spectra to the normalized `mass_dataset` before feature-network detection.

```{r ms2-summary}
if (nrow(ms2_summary) > 0) {
  knitr::kable(ms2_summary, caption = "MS2 files and feature-level MS2 attachment summary.")
} else {
  cat("No MS2 attachment summary was found.")
}
```

## Feature Network

::: {.callout-note appearance="simple" icon="false"}
## What Is a Feature Network?

In LC-MS data, one real metabolite often appears as several instrument signals instead of one clean peak.
These extra signals can come from natural isotopes, different charged forms, salt adducts, fragments made inside the instrument, or the same metabolite observed in positive and negative modes.
The Feature Network connects signals that probably come from the same original molecule, so MetMiner can avoid counting the same metabolite many times and can choose a cleaner representative annotation for downstream analysis.

For the full technical rationale, see the [Feature Network principles document](https://github.com/xuebinzhang-lab/MetMiner/blob/feature/global-ai-bot/inst/app/feature_network_principles.md).
:::

Feature-network evidence is used to support recurrent ion inspection, subnetwork roles, pseudo-area aggregation, and downstream redundancy filtering.

```{r feature-network-summary}
if (nrow(network_summary) > 0) {
  knitr::kable(network_summary, caption = "Feature-network summary by polarity.")
} else {
  cat("No feature-network summary was found.")
}
```

```{r feature-network-plot}
if (nrow(network_summary) > 0 && "mode" %in% colnames(network_summary)) {
  net <- network_summary |>
    filter(mode %in% c("positive", "negative")) |>
    mutate(across(c(n_features, n_edges, n_role_rows, n_pseudo_compounds), as.numeric)) |>
    pivot_longer(c(n_features, n_edges, n_role_rows, n_pseudo_compounds), names_to = "metric", values_to = "value")
  p <- ggplot(net, aes(metric, value, fill = mode)) +
    geom_col(position = "dodge", width = 0.72) +
    labs(x = NULL, y = "Count", title = "Feature-Network Outputs") +
    theme_light() +
    theme(axis.text.x = element_text(angle = 20, hjust = 1))
  plotly_or_static(p, max(net$value, na.rm = TRUE))
}
```

::: {.panel-tabset}

## Positive Network

```{r feature-network-positive-widget}
feature_network_widget(feature_network_pos, "Positive Feature Network", max_edges = 700)
```

## Negative Network

```{r feature-network-negative-widget}
feature_network_widget(feature_network_neg, "Negative Feature Network", max_edges = 700)
```

## Merged Polarity Network

```{r feature-network-merged-widget}
feature_network_widget(feature_network_merged, "Merged Positive/Negative Feature Network", max_edges = 900)
```

:::

::: {.callout-tip title="Network Legend" collapse="true"}
The interactive feature-network widgets display feature relationships discovered before annotation filtering.
Nodes are LC-MS features, while edges represent isotope, adduct, in-source fragment, recurrent, or cross-polarity relationships depending on the detected relation type.
The report displays the highest-scoring subset of large networks to keep the HTML responsive; the complete network is retained in the saved `06.*feature_network*.rda` objects.
:::

## Database Annotation

Annotation combines species/resource-aware plant databases with public MS2 libraries.
QC samples are removed before annotation so the biological annotation objects contain only experimental samples.

```{r annotation-summary}
if (nrow(annotation_summary) > 0) {
  knitr::kable(annotation_summary, caption = "Database annotation summary.")
} else {
  cat("No annotation summary was found.")
}
```

```{r annotation-plot}
if (nrow(annotation_summary) > 0) {
  ann <- annotation_summary |>
    mutate(across(c(raw_annotation_rows, display_annotation_rows, annotated_features, removed_qc_count), as.numeric)) |>
    select(mode, raw_annotation_rows, annotated_features, removed_qc_count) |>
    pivot_longer(-mode, names_to = "metric", values_to = "value")
  p <- ggplot(ann, aes(metric, value, fill = mode)) +
    geom_col(position = "dodge", width = 0.72) +
    labs(x = NULL, y = "Count", title = "Annotation Candidates and QC Removal") +
    theme_light() +
    theme(axis.text.x = element_text(angle = 20, hjust = 1))
  plotly_or_static(p, max(ann$value, na.rm = TRUE))
}
```

## Non-Redundant Filtering

The filtering layer prioritizes annotation confidence, feature-network support, adduct evidence, same-compound redundancy, recurrent ion context, and cross-polarity duplication.

```{r filtering-summary}
if (nrow(filter_summary) > 0) {
  knitr::kable(filter_summary, caption = "Annotation filtering and redundancy-removal summary.")
} else {
  cat("No annotation filtering summary was found.")
}
```

::: {.panel-tabset}

## Final Non-Redundant

```{r final-preview}
if (nrow(final_annotations) > 0) {
  cols <- intersect(c("mode", "metabolite_id", "compound_name", "annotation_level", "annotation_layer", "record_type", "representative_feature", "n_features", "rt", "mz"), colnames(final_annotations))
  knitr::kable(utils::head(final_annotations[, cols, drop = FALSE], 20), caption = "Top rows from the final non-redundant annotation table.")
} else {
  cat("No final annotation table was found.")
}
```

## Collapse Review

```{r collapse-preview}
if (nrow(collapse_annotations) > 0) {
  knitr::kable(utils::head(collapse_annotations, 20), caption = "Top rows from the collapsed review table.")
} else {
  cat("No collapse annotation table was found.")
}
```

## Expand Review

```{r expand-preview}
if (nrow(expand_annotations) > 0) {
  knitr::kable(utils::head(expand_annotations, 20), caption = "Top rows from the expanded review table.")
} else {
  cat("No expand annotation table was found.")
}
```

:::

------------------------------------------------------------------------

# ClassyFire Classification

The classification summary is calculated from the final clean non-redundant annotations.
Missing classes are filled from the original annotated objects when possible and reported as `Unclassified` otherwise.

```{r classification-status}
if (nrow(classification_status) > 0) {
  knitr::kable(classification_status, caption = "Classification result status.")
}
```

::: {.panel-tabset}

## Super Class

```{r super-class-pie}
pie_plot(classification_summary, "Super_class")
```

```{r super-class-table}
knitr::kable(
  classification_detail_table("Super_class"),
  caption = "Final annotations with Super_class, Class, Sub_class, and available parent levels."
)
```

## Class

```{r class-pie}
pie_plot(classification_summary, "Class")
```

```{r class-table}
knitr::kable(
  classification_detail_table("Class"),
  caption = "Final annotations with Super_class, Class, Sub_class, and available parent levels."
)
```

## Sub Class

```{r subclass-pie}
pie_plot(classification_summary, "Sub_class")
```

```{r subclass-table}
knitr::kable(
  classification_detail_table("Sub_class"),
  caption = "Final annotations with Super_class, Class, Sub_class, and available parent levels."
)
```

:::

::: {.callout-tip title="Figure Legend" collapse="true"}
Pie charts summarize the final non-redundant annotation table across ClassyFire Super_class, Class, and Sub_class dimensions.
Categories beyond the top 12 are grouped as `Other`.
Pie charts are rendered as interactive plotly charts.
:::

------------------------------------------------------------------------

# Output Tables

```{r output-table-list}
outputs <- data.frame(
  output = c(
    "08.final_annotations.tsv",
    "08.expand_annotations.tsv",
    "08.collapse_annotations.tsv",
    "08.redundancy_audit.tsv",
    "09.classification_records.tsv",
    "09.classification_summary.tsv"
  ),
  rows = c(
    nrow(final_annotations),
    nrow(expand_annotations),
    nrow(collapse_annotations),
    nrow(redundancy_audit),
    nrow(classification_records),
    nrow(classification_summary)
  ),
  stringsAsFactors = FALSE
)
knitr::kable(outputs, caption = "Downloadable annotation and classification tables.")
```

<details>
<summary><strong>Pipeline Status Objects</strong></summary>

```{r status-objects}
status_overview <- data.frame(
  step = c("annotation", "annotation_filtering", "classification"),
  state = c(annotation_status$state %||% NA_character_, filter_status$state %||% NA_character_, classification_json$state %||% NA_character_),
  updated_at = c(annotation_status$updated_at %||% NA_character_, filter_status$updated_at %||% NA_character_, classification_json$updated_at %||% NA_character_),
  stringsAsFactors = FALSE
)
knitr::kable(status_overview, caption = "Pipeline status metadata.")
```

</details>
