---
title: "Data Quality Assessment Report"
date: last-modified
format: 
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    self-contained: true
    code-tools: false
    smooth-scroll: true
params:
  work_dir: null
  viz: null
editor: 
  markdown: 
    wrap: sentence
---


```{r setup, include=FALSE}
#| label: setup
#| include: false

# Fix: Remove duplicate {r setup, include=FALSE}, keep only #| options
knitr::opts_chunk$set(
  echo = FALSE, 
  warning = FALSE, 
  message = FALSE, 
  fig.align = "center", 
  fig.width = 10, 
  fig.height = 6
)

library(massdataset)
library(massqc)
library(ggplot2)
library(dplyr)
library(plotly)

# --- LOAD DATA FROM PARENT ENVIRONMENT ---
# Load objects directly from parent environment to avoid file path issues
object_pos <- NULL
object_neg <- NULL

# Try loading objects from parent environment
if (!is.null(params$work_dir)) {
  pos_path <- file.path(params$work_dir, "object_pos.rds")
  neg_path <- file.path(params$work_dir, "object_neg.rds")
  
  if (file.exists(pos_path)) {
    object_pos <- readRDS(pos_path)
  }
  if (file.exists(neg_path)) {
    object_neg <- readRDS(neg_path)
  }
}

# Helper for preprocessing (Log + Scale for PCA/Boxplot)
preprocess_data <- function(obj, scale = FALSE) {
  if (is.null(obj)) return(NULL)
  temp <- obj
  # Log transformation (obj + 1 first to handle zeros)
  temp <- log(temp + 1, 2)
  
  if(scale) {
    # Manual Scaling on expression_data (Z-score)
    ed <- temp@expression_data
    # Scale operates on columns, so transpose -> scale -> transpose back
    scaled <- t(scale(t(ed)))
    temp@expression_data <- as.data.frame(scaled)
  }
  return(temp)
}
```

::: {.callout-note icon="false" appearance="simple"}
## Report Information

**Software**: [MetMiner by Wang et al., 2024](https://github.com/ShawnWx2019/MetMiner), powered by [TidyMass2 by Wang et al., 2026](https://www.nature.com/articles/s41467-026-68464-7)

**Date**: `r format(Sys.Date(), '%Y-%m-%d')`

**Generated By**: [MetMiner-shinyapp](https://github.com/ShawnWx2019/MetMiner)
:::

------------------------------------------------------------------------

# Introduction

This report provides a comprehensive quality assessment of your metabolomics data.
It is generated by **MetMiner**, a user-friendly pipeline for large-scale plant metabolomics data analysis, leveraging the power of **TidyMass2**.

------------------------------------------------------------------------

# Sample Information

::::: grid
::: g-col-6
```{r}
#| results: asis

# NOTE: Use 'object_pos' directly, NOT 'params$object_pos'
if(!is.null(object_pos)){
  cat("### Positive Mode\n")
  cat(paste0("- **Samples**: ", ncol(object_pos), "\n"))
  cat(paste0("- **Variables**: ", nrow(object_pos), "\n"))
  cat(paste0("- **Missing Values**: ", sum(is.na(object_pos@expression_data)), "\n"))
} else {
  cat("### Positive Mode\n*No data loaded*")
}
```
:::

::: g-col-6
```{r}
#| results: asis

# NOTE: Use 'object_neg' directly
if(!is.null(object_neg)){
  cat("### Negative Mode\n")
  cat(paste0("- **Samples**: ", ncol(object_neg), "\n"))
  cat(paste0("- **Variables**: ", nrow(object_neg), "\n"))
  cat(paste0("- **Missing Values**: ", sum(is.na(object_neg@expression_data)), "\n"))
} else {
  cat("### Negative Mode\n*No data loaded*")
}
```
:::
:::::

------------------------------------------------------------------------

# Quality Control Plots

## m/z vs Retention Time

::: panel-tabset
### Positive Mode

```{r}
#| fig-height: 5

if(!is.null(object_pos)){
  p1 <- massdataset::show_mz_rt_plot(object_pos, hex = params$viz$mzrt_hex) + 
    ggtitle("Positive Mode") + theme_light()
  print(p1)
} else {
  cat("No Positive Mode Data")
}
```

### Negative Mode

```{r}
#| fig-height: 5

if(!is.null(object_neg)){
  p2 <- massdataset::show_mz_rt_plot(object_neg, hex = params$viz$mzrt_hex) + 
    ggtitle("Negative Mode") + theme_light()
  print(p2)
} else {
  cat("No Negative Mode Data")
}
```
:::

::: {.callout-tip title="Figure Legend" collapse="true"}
The scatter plot (or hexbin plot) displays the distribution of all detected metabolic features across **Retention Time (RT)** and **Mass-to-Charge Ratio (m/z)**.  

- **Uniform distribution** is generally expected for untargeted metabolomics.
:::

------------------------------------------------------------------------

## Sample Missing Values

::: panel-tabset
### Positive Mode

```{r}
#| fig-height: 6

viz <- params$viz
if(!is.null(object_pos)){
  p <- massdataset::show_sample_missing_values(
    object = object_pos,
    color_by = viz$mv_color,
    order_by = viz$mv_order,
    percentage = viz$mv_pct,
    show_x_text = viz$mv_text,
    desc = viz$mv_desc
  ) + ggtitle("Positive: Missing Values") + theme_light() + theme(axis.text.x = element_text(angle=45, hjust=1))
  print(p)
}
```

### Negative Mode

```{r}
#| fig-height: 6

if(!is.null(object_neg)){
  p <- massdataset::show_sample_missing_values(
    object = object_neg,
    color_by = viz$mv_color,
    order_by = viz$mv_order,
    percentage = viz$mv_pct,
    show_x_text = viz$mv_text,
    desc = viz$mv_desc
  ) + ggtitle("Negative: Missing Values") + theme_light() + theme(axis.text.x = element_text(angle=45, hjust=1))
  print(p)
}
```
:::

::: {.callout-tip title="Figure Legend" collapse="true"}
Each point represents a sample.

The Y-axis indicates the **Missing Value (NA)** count or percentage.

\- **Color**: `r viz$mv_color`

\- **High missing values** in QC samples may indicate instrument instability.
:::

------------------------------------------------------------------------

## PCA Analysis

::: panel-tabset
### Positive Mode

```{r}
#| fig-height: 6

viz <- params$viz
do_pca <- function(obj, title) {
  temp_obj <- preprocess_data(obj, scale = viz$pca_scale)
  p <- massqc::massqc_pca(
    object = temp_obj,
    color_by = viz$pca_color,
    order_by = "sample_id",
    point_alpha = viz$pca_alpha,
    point_size = viz$pca_size
  ) + ggtitle(title) + theme_light()
  print(p)
}

if(!is.null(object_pos)) do_pca(object_pos, "Positive Mode PCA")
```

### Negative Mode

```{r}
#| fig-height: 6

if(!is.null(object_neg)) do_pca(object_neg, "Negative Mode PCA")
```
:::

::: {.callout-tip title="Figure Legend" collapse="true"}
Principal Component Analysis (PCA) score plot of PC1 vs PC2.

\- **Preprocessing**: Log2 transformed `r if(viz$pca_scale) "and Z-score Scaled"`.

\- **Interpretation**: QC samples should cluster tightly.
Separate clusters may indicate batch effects or biological differences.
:::

------------------------------------------------------------------------

## Cumulative RSD

::: panel-tabset
### Positive Mode

```{r}
#| fig-height: 5

viz <- params$viz
if(!is.null(object_pos)){
  p <- massqc::massqc_cumulative_rsd_plot(
    object_pos, 
    rsd_cutoff = viz$rsd_cutoff,
    color = "black"
  ) + ggtitle(paste0("Positive Mode (Cutoff: ", viz$rsd_cutoff, "%)")) + theme_light()
  print(p)
}
```

### Negative Mode

```{r}
#| fig-height: 5

if(!is.null(object_neg)){
  p <- massqc::massqc_cumulative_rsd_plot(
    object_neg, 
    rsd_cutoff = viz$rsd_cutoff,
    color = "black"
  ) + ggtitle(paste0("Negative Mode (Cutoff: ", viz$rsd_cutoff, "%)")) + theme_light()
  print(p)
}
```
:::

::: {.callout-tip title="Figure Legend" collapse="true"}
Cumulative distribution of Relative Standard Deviation (RSD).

\- **Y-axis**: Percentage of features with RSD \< X.

\- **Goal**: A steep curve rising early indicates high precision (most features have low RSD).
:::

------------------------------------------------------------------------

## Sample Boxplot & Correlation

::: panel-tabset
### Boxplot (Pos)

```{r}
#| fig-height: 5

viz <- params$viz
do_boxplot <- function(obj, title) {
  temp_obj <- log(obj + 1, 2)
  p <- massqc::massqc_sample_boxplot(
    object = temp_obj,
    color_by = viz$box_color,
    fill_by = viz$box_fill,
    point = viz$box_point
  ) + ggtitle(title) + ylab("Log2 Intensity") + theme_light() + theme(axis.text.x = element_text(angle=45, hjust=1))
  print(p)
}
if(!is.null(object_pos)) do_boxplot(object_pos, "Positive Mode")
```

### Boxplot (Neg)

```{r}
#| fig-height: 5

if(!is.null(object_neg)) do_boxplot(object_neg, "Negative Mode")
```

### Correlation (Pos)

```{r}
#| fig-height: 7

do_cor <- function(obj, title) {
  ord <- if(viz$cor_order == "no") NULL else viz$cor_order
  p <- massqc::massqc_sample_correlation(
    object = obj,
    cor_method = viz$cor_method,
    order_by = ord
  ) + ggtitle(title)
  print(p)
}
if(!is.null(object_pos)) do_cor(object_pos, "Positive Mode")
```

### Correlation (Neg)

```{r}
#| fig-height: 7

if(!is.null(object_neg)) do_cor(object_neg, "Negative Mode")
```
:::

------------------------------------------------------------------------

# Citation

Please cite the following papers if you use MetMiner in your research:

> **MetMiner**: Wang, X., Liang, S., Yang, W., Yu, K., Liang, F., Zhao, B., Zhu, X., Zhou, C., Mur, L.A.J., Roberts, J.A., Zhang, J., Zhang, X., 2024.
> MetMiner: A user-friendly pipeline for large-scale plant metabolomics data analysis.
> *J. Integr. Plant Biol.* 66, 2329–2345.
> [doi:10.1111/jipb.13774](https://doi.org/10.1111/jipb.13774)

> **TidyMass2**: Wang, X., Liu, Y., Jiang, C., Huang, Z., Yan, H., Wong, S.H., Johnson, C.H., Zhang, J., Ge, Y., Zhang, F., Zhang, J., Lai, R., Gao, P., Zhang, X., Shen, X., 2026.
> TidyMass2: advancing LC-MS untargeted metabolomics through metabolite origin inference and metabolic feature-based functional module analysis.
> *Nat. Commun.* 17, 1755.
> [doi:10.1038/s41467-026-68464-7](https://doi.org/10.1038/s41467-026-68464-7)

<br> <br>

::: {.column-screen-inset style="background-color: #f8f9fa; padding: 20px; text-align: center; border-top: 1px solid #dee2e6;"}
<small style="color: #6c757d;"> © 2024-2025 **MetMiner** \| State Key Laboratory of Crop Stress Adaptation and Improvement<br> Henan University, Kaifeng 475004, China </small>
:::
